---
title: "Bayesian Reconstruction of Partially Observed Tongue Shapes"
author: "Katrin Kulberg, Louis Arthur David Del Perugia, Salomé Olivier Moret, Maximilian Brunner, Cheikh Mourtada Mboup"
date: "11.05.2026"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# libraries
library(dplyr)
library(tidyr)
library(MASS)
library(Matrix)
library(ggplot2)
library(ggforce) # (extra plotting features)
library(ggpattern) # (extra plotting features)
library(fitdistrplus)
library(gridExtra)
#library(grid) 

# loading data (assuming the files are in the same folder as the Rmd file)
load("speaker1_tongue_shapes.RData")
load("speaker2_tongue_shapes.RData")
```

# Data Preprocessing

This subsection takes care of the initial data processing, including creating dataframe objects for the data of both speakers, reordering the points of the second speaker, and creating the standardised coordinate ($\tau$) along the curve.

```{r}
# transforming data into a data frame shape
# first speaker
df1 <- data.frame(
  x = as.vector(speaker1[,1,]),
  y = as.vector(speaker1[,2,]),
  t = rep(1:dim(speaker1)[1], times = dim(speaker1)[3]),     # point along the curve
  shape = rep(1:dim(speaker1)[3], each = dim(speaker1)[1]),  # shape number
  speaker = 1
)

# second speaker
df2 <- data.frame(
  x = as.vector(speaker2[,1,]),
  y = as.vector(speaker2[,2,]),
  t = rep(1:dim(speaker2)[1], times = dim(speaker2)[3]),     # point along the curve
  shape = rep(1:dim(speaker2)[3], each = dim(speaker2)[1]),  # shape number
  speaker = 2
)

# given that the curves of the second speaker appear to be indexed in an opposite order compared to the curves of the first speaker, we reorder the indices t for speaker 2
df2$t <- 45 - df2$t + 1

# however, given that the 5th shape was already in the right order, we reorder that particular shape again
# additionally removing an outlier from shape 5
df2 <- df2 %>%
  mutate(t= ifelse(shape == 5, 45 - t + 1, t)) %>% 
  filter(!(shape == 5 & t == 45)) # remove outlier


# adding standardised index t (tau on the poster) to both data frames (this step allows to combine curves from speaker 1 and 2 in the later stage)
# (multiplying by 1/(n+1)), where n is the number of data points
df1$t_standardised <- df1$t / 66
df2$t_standardised <- df2$t / 46

# combining data frames of both speaker
df <- rbind(df1, df2)
```


# General Function Definitions

This subsection includes general functions that will be used in a later stage.

```{r}

# function that constructs the normalised shifted Legendre basis for input x and degree
# function is only implemented up to degree 5, as it is the highest degree used in this project
build_shifted_legendre_basis <- function(x, degree) {
  
  # binding column-wise
  sapply(0:degree, function(n) {
    
    Pn <- if (n == 0) {
      rep(1, length(x))
      
    } else if (n == 1) {
      2*x - 1
      
    } else if (n == 2) {
      6*x^2 - 6*x + 1
      
    } else if (n == 3) {
      20*x^3 - 30*x^2 + 12*x - 1
      
    } else if (n == 4) {
      70*x^4 - 140*x^3 + 90*x^2 - 20*x + 1
      
    } else if (n == 5) {
      252*x^5 - 630*x^4 + 560*x^3 - 210*x^2 + 30*x - 1
      
    }
    
    # normalisation step
    sqrt(2*n + 1) * Pn
  })}


# function that computes least-squares coefficients in the normalised shifted Legendre basis up to a given degree
coefs_shifted_legendre <- function(x, y, degree){
  
  X_shifted_legendre <- build_shifted_legendre_basis(x, degree)
  model <- lm(y ~ X_shifted_legendre - 1)
  beta <- coef(model)
  
  return(beta)}


# function that fits a linear model in the normalised shifted Legendre basis and returns the estimated residual variance (sigma^2)
calculate_sigma2 <- function(x, y, degree){
  
  X_poly <- build_shifted_legendre_basis(x, (degree))
  fitted_model <- lm(y ~ X_poly - 1)
  
  return(summary(fitted_model)$sigma^2)
}


# function that computes normalised shifted Legendre basis coefficients for each curve and returns a matrix where each row corresponds to a shape
# data frame, name of the response variable ("x" or "y"), degree of the polynomial, and vector of shape identifiers (with default being all fully observed curves) to include can be specified
compute_betas_shifted_legendre <- function(df, variable, degree, shapes = 3:10){
  
  # collecting into a list
  betas_list <- lapply(shapes, function(i) {
    
    # selecting response (x or y) from a given data frame
    y <- df[df$shape == i, variable]
    
    # calculating coefficients based on standardised index t (tau)
    coefs_shifted_legendre(df[df$shape == i, "t_standardised"], y, degree)
  })
  
  do.call(rbind, betas_list)
}


```

# Plotting Functions

This subsection includes functions that will be later used to illustrate the results graphically.

```{r}

# function for plotting reconstructed marginal curves (x or y)
# both the best estimate and all possible trajectories are plotted
# it is possible to specify through the parameter "mode" whether to plot trajectories for the whole curve or only for the unknown points, with the former being the default
summarise_results_1d <- function(predictions, points, true_data, non_observed, response, mode = "full"){
  
  # if mode is "full", trajectories will be plotted throughout the curve
  if(mode == "full"){
    non_observed <- 1:length(points)
  }
  
  # building a summarising data frame (best estimate)
  df <- data.frame(
    best_estimate = apply(predictions, 1, quantile, probs = 0.5))
  
  # number of all posterior trajectories
  n_samples <- ncol(predictions)
  
  # building list of all trajectories
  list_of_trajectories <- lapply(1:n_samples, function(j) {
    data.frame(
      sample = j,
      points = points[non_observed],
      values = predictions[non_observed, j]
    )
  })

  # combining into a single data frame
  df_trajectories <- do.call(rbind, list_of_trajectories)

  
  # plotting
  p <- ggplot() + 
    geom_line(data = df_trajectories, aes(x = points,y = values, group = sample), colour = "grey50", alpha = 0.05, linewidth = 0.1) + 
    geom_line(data = df, aes(x = points, y = best_estimate), colour = "black", linewidth = 1) +
    geom_point(aes(x = points, y = true_data), colour = "red") + labs(x = expression(tau), y = response) + theme_bw()
  
  
  return(p)
  
}


# function for plotting the reconstructed curve (x, y)
# both the best estimate and all possible trajectories are plotted
# it is possible to specify through the parameter "mode" whether to plot trajectories for the whole curve or only for the unknown points, with the former being the default
summarise_results_2d <- function(predictions_x, predictions_y, 
                                true_data_x, true_data_y, non_observed, xlim, ylim, mode = "full"){
  
  
  # if mode is "full", trajectories will be plotted throughout the curve
  if(mode == "full"){
    non_observed <- 1:nrow(predictions_x)
  }
  
  # building a summarising data frame (best estimate)
  df_xy <- data.frame(
  best_estimate_x = apply(predictions_x, 1, quantile, probs = 0.5),
  best_estimate_y = apply(predictions_y, 1, quantile, probs = 0.5))
  
  # number of all posterior trajectories
  n_samples <- ncol(predictions_x)
  
  # building list of all trajectories
  list_of_trajectories <- lapply(1:n_samples, function(j) {
    data.frame(
      sample = j,
      x = predictions_x[non_observed, j],
      y = predictions_y[non_observed, j]
    )
  })
  

  # binding into a data frame for plotting
  df_trajectories <- do.call(rbind, list_of_trajectories)
  
  # plotting
  p <- ggplot() +
    geom_path(data = df_trajectories, aes(x = x, y = y, group = sample), colour = "grey50", alpha = 0.05, linewidth = 0.1) +
    geom_path(data = df_xy, aes(x = best_estimate_x, y = best_estimate_y), colour = "black", linewidth = 1.25) +
    geom_point(aes(x = true_data_x, y = true_data_y), colour = "red", size = 2) +
    labs(x = NULL, y = NULL) + coord_cartesian(xlim = xlim, ylim = ylim)+ theme_bw()
  
  return(p)
  
}

################################################################################
################################################################################
# the following includes functions for plotting results in case of joint modelling

# function that computes density thresholds defining 2D HPD regions for given probability levels based on kernel density estimation
# retrieved from https://stackoverflow.com/questions/70569213/two-parameter-bayesian-credibility-interval
get_hpd2d_levels <- function(x, y, prob=c(0.9,0.95), ...) {
  post1 <- MASS::kde2d(x, y)
  dx <- diff(post1$x[1:2])
  dy <- diff(post1$y[1:2])
  sz <- sort(post1$z)
  c1 <- cumsum(sz) * dx * dy
  ## remove duplicates
  ## dups <- duplicated(sz)
  ## sz <- sz[!dups]
  ## c1 <- c1[!dups]
  levels <- sapply(prob, function(x) {
    approx(c1, sz, xout = 1 - x, ties = mean)$y
  })
  return(levels)
}

# function for plotting the reconstructed curve (x, y) in case of joint modelling 
# credible region and the true data are also portrayed
# it is possible to specify through the parameter "mode" whether to plot credible region for the whole curve or only for the unknown points, with the latter being the default
# under joint modelling, x and y are no longer assumed to be independent
# credible regions are obtained using highest posterior density (HPD) contours from kernel density estimation
summarise_results_2d_joint <- function(predictions_x, predictions_y, points, 
                                       true_data_x, true_data_y, non_observed, mode = "only_missing"){
  
  # building a summarising data frame (best estimate) for both x and y
  df_xy <- data.frame(
    best_estimate_x = apply(predictions_x, 1, quantile, probs = 0.5),
    best_estimate_y = apply(predictions_y, 1, quantile, probs = 0.5))
  
  
  # if only the missing region is requested, ellipses will be drawn only for that region
  if(mode == "only_missing"){
    indices <- non_observed
  } else{
    indices <- seq_len(nrow(predictions_x))
  }
  
  # initalising plot
  p <- ggplot()
  
  for(i in indices) {
    
    # density threshold corresponding to the 95% HPD region
    L <- get_hpd2d_levels(predictions_x[i,], predictions_y[i,], prob = 0.95)
    # kernel density estimation
    kernel_density <- kde2d(predictions_x[i,], predictions_y[i,])
    # contour lines at the hpd level
    cl <- contourLines(kernel_density, levels = L)
    # combining into data frame for plotting
    df_contour <- do.call(rbind, lapply(cl, function(c) data.frame(x = c$x, y = c$y)))
    # plotting 95% credible region for a given point
    p <- p + geom_polygon(data = df_contour, aes(x = x, y = y), fill = "grey50", alpha = 0.3)
  }
  
  # adding remaining parts(best estimates and existing data)
  p <- p +
    geom_path(data = df_xy, aes(x = best_estimate_x, y = best_estimate_y), linewidth = 1.25) +
    geom_point(aes(x = true_data_x, y = true_data_y), color = "red", size = 2) + 
    labs(x = NULL, y = NULL) + theme_bw()
  
  return(p)
  
}


################################################################################
# the following includes functions for plotting results from multiple marginal modelling methods on same figure

# function that constructs a data frame containing marginal model results needed for plotting
build_df_1d <- function(predictions, points, non_observed, mode = "full", model_name){
  
  
  # if mode is "full", trajectories will be plotted throughout the curve
  if(mode == "full"){
    non_observed <- 1:length(points)
  }
  
  # number of all posterior trajectories
  n_samples <- ncol(predictions)
  
  
  # building list of all trajectories
  list_of_trajectories <- lapply(1:n_samples, function(j) {
    data.frame(
      sample = j,
      t = points[non_observed],
      value = predictions[non_observed, j],
      model = model_name )})

  # binding into a data frame
  df_trajectories <- do.call(rbind, list_of_trajectories)
  # unique id
  df_trajectories$group_id <- paste(model_name, df_trajectories$sample, sep = "_")
  
  
  # best estimate
  df_best <- data.frame(
    t = points,
    best_estimate = apply(predictions, 1, median),
    model = model_name)
  
  
  # returning both data frames
  return(list(
    trajectories = df_trajectories,
    best = df_best
  ))
}



# function that plots multiple marginal modelling methods on same figure
plot_1d_models <- function(model_list, points, true_data, non_observed, colors, response, model_order, mode = "full"){
  
  # initialising a list for models
  df_list <- list()

  # looping over models; each time saving the data
  for(name in model_order){
    
    
    # this will return all trajectories and the best estimates
    res <- build_df_1d(
      predictions = model_list[[name]]$predictions,
      points = points,
      non_observed = non_observed,
      mode = mode,
      model_name = name
    )

    df_list[[name]] <- res
  }
  

  # binding separately
  # all trajectories from data frame df_trajectories
  # all best estimates from data frame df_best
  df_trajectory <- do.call(rbind, lapply(df_list, `[[`, "trajectories"))
  df_best <- do.call(rbind, lapply(df_list, `[[`, "best"))

  # reorder for plotting
  df_trajectory$model <- factor(df_trajectory$model, levels = model_order)
  df_best$model <- factor(df_best$model, levels = model_order)

  # data frame for true data
  df_true <- data.frame(t = points, y = true_data)

  # plotting
  p <- ggplot() + 
    geom_path(data = df_trajectory, aes(x = t, y = value, group = group_id, color = model), 
    alpha = 0.01, linewidth = 0.1) +
    geom_path(data = df_best, aes(x = t, y = best_estimate, color = model), linewidth = 1) +
    geom_point(data = df_true, aes(x = t, y = y), color = "black", size = 1) +
    scale_color_manual(values = colors) + labs(x = expression(tau), y = response) +
    guides(color = "none", pattern_colour = "none", pattern_angle = "none") + theme_bw()
  
  return(p)
}


################################################################################
################################################################################

# function that returns a collection of trace plots for each coefficient with the help of gridExtra 
# exact layout of the plot can be specified
summarise_trace_plots <- function(beta_samples, plot_rows, plot_cols){
  
  # list of ggplot objects
  p <- ncol(beta_samples)
  plot_list <- lapply(1:p, function(j){
  # extracting data for a given coefficient
  df <- data.frame(
    number = 1:nrow(beta_samples),
    values = beta_samples[, j]
  )
  # trace plots
  ggplot(df, aes(x = number, y = values)) +
    geom_line() +
    labs(title = NULL, x = bquote(beta[.(j)]), y = "Value") + theme_bw()
})
  return(grid.arrange(grobs = plot_list, nrow = plot_rows, ncol = plot_cols))
  
}



```


# Modelling

## Determining an Appropriate Prior for $\sigma^2$

In order to set reasonable parameters for the inverse-gamma prior on $\sigma^2$, we estimate the empirical distribution of $\sigma^2$ from fully observed curves and determine the corresponding $\nu_0$ and $\sigma_{0}^2$.

```{r}

# we are interested in finding estimated sigmas for all the fully observed curves of both speakers 
curves_to_check <- expand.grid(3:10, c(1, 2))
# initialising data structure to store values
sigmas <- matrix(NA, nrow = 16, ncol = 2)

for (i in 1:nrow(curves_to_check)) {
  
  # selecting combination
  pair <- curves_to_check[i, ]
  
  shape_i <- as.numeric(pair[1])
  speaker_i <- as.numeric(pair[2])
  
  # extracting the data
  this_curve <- subset(df, df$speaker == speaker_i & df$shape == shape_i)
  
  # calculating estimate of sigma^2 for x and y
  sigmas[i,1] <- calculate_sigma2(this_curve$t_standardised, this_curve$x, 5)
  sigmas[i,2] <- calculate_sigma2(this_curve$t_standardised, this_curve$y, 4)
  
}


# finding the gamma parameters for distribution of 1/sigma^2
gamma <- fitdist(as.vector((1/sigmas)), "gamma", lower=c(0,0))
shape <- gamma$estimate[1]
rate <- gamma$estimate[2]
# as a reference:
# shape = 1.59865
# rate = 0.3105013  

# checking the fit graphically
hist(as.vector((1/sigmas)), probability = TRUE)
curve(dgamma(x, shape = shape, rate = rate), add = TRUE, col = "red", lwd = 1)
# => seems reasonable
  
# determining parameters for the prior
nu <- 2 * shape
sigma2 <- (2*rate) / (nu)

```

## Gibbs Sampling Procedures

The following subsection contains all functions related to Gibbs Sampling.

```{r}
# global parameters
default_nu0 <- nu
default_sigmasq_0 <- sigma2
default_first_sigma <- 2 # initial sigma^2 in Gibbs Sampling

################################################################################

# function that selects the right number of curves based on the observed data
# by default, ceiling((degree + 1)/2) best-fitting and floor((degree + 1)/2) worst-fitting curves are selected
curve_selection <- function(xs, ys, prediction_interval, coefs, number = -1){

  num_select <- number
  
  # by default, using the total number of independent polynomials, which corresponds to the number of
  # columns in the coefficient matrix (degree + 1)
  if (number <= 0) {
    num_select <- ncol(coefs)
  }
  
  # determining the missing and observed regions
  missing_idx = (xs > prediction_interval[1]) & (xs < prediction_interval[2])
  observed_idx = !missing_idx
  
  # filtering out the observed part
  Y_observed <- ys[observed_idx]
  
  # building basis (with normalised shifted Legendre polynomials)
  X_poly_observed <- build_shifted_legendre_basis(xs[observed_idx], (ncol(coefs)-1))
  
  # constructing the design matrix (evaluating the polynomials)
  X_observed <- t((coefs) %*% t(X_poly_observed))
  
  # initialising data structure to store values
  results <- matrix(NA, nrow = nrow(coefs), ncol = 2)
  
  for (i in 1:ncol(X_observed)){
    
    # computing similarity (based on squared difference)
    similarity <- sum((X_observed[,i] - Y_observed)^2)
    results[i,1] <- i
    results[i,2] <- similarity
  }
  
  # sorting by similarity in an ascending order
  results <- results[order(results[,2]), ]
  
  # (a %/% b == floor(a/b) for integers)
  good_number <- num_select %/% 2 + (num_select %% 2 != 0)
  bad_number  <- num_select %/% 2
  
  return(coefs[results[c(1:good_number, (nrow(results) - bad_number):nrow(results)),1], ])
  
}


# function that performs Gibbs Sampling assuming that required quantities have been precomputed
Gibbs_core <- function(t, y, shapes_to_fit, prediction_interval,
                       beta0, Sigma_0, nu0, sigmasq_0,
                       first_beta, first_sigma, n_simulations = 5000, n_burnin = 100) {
	
  
  num_params <- ncol(shapes_to_fit)
	n <- length(y)
	
	# a vector that has TRUE for all entries that are missing
	missing_idx <- (t > prediction_interval[1]) & (t < prediction_interval[2])
	
	# a vector that has TRUE for all entries that are observed
	observed_idx <- !missing_idx
	
  # selecting only observed part of curve (for both design matrix and response y)
	observed_shapes <- shapes_to_fit[observed_idx,]
	y_observed <- y[observed_idx]
	
	Sigma0_inv <- solve(Sigma_0) # for later use
	
  # initialising data structures to store values
	beta_samples <- matrix(NA, nrow=(n_simulations + n_burnin), ncol = num_params)
	sigma2_samples <- numeric((n_simulations + n_burnin))
	
  # setting initial values
	beta_samples[1, ] <- first_beta
	sigma2_samples[1] <- first_sigma
	
	for(iter in 2:(n_simulations + n_burnin)){
		
		# UPDATING BETAS
		
		# computing covariance matrix
		V_beta <- solve(t(observed_shapes) %*% observed_shapes / sigma2_samples[iter-1] + Sigma0_inv)
		# computing mean vector
		m_beta <- V_beta %*% (t(observed_shapes) %*% y_observed / sigma2_samples[iter-1] + Sigma0_inv %*% beta0)
		
		# sampling beta
		beta_samples[iter, ] <- mvrnorm(1, mu = m_beta, Sigma = V_beta)
		
		# UPDATING SIGMA^2
		
		# computing SSR
		SSR <- sum((y_observed - observed_shapes %*% beta_samples[iter, ])^2)

		# computing parameters for inverse gamma
		shape_sig <- (nu0 + n) / 2
		rate_sig  <- (nu0 * sigmasq_0 + SSR) / 2
		
		# sampling sigma^2
		sigma2_samples[iter] <- 1 / rgamma(1, shape = shape_sig, rate = rate_sig)
	}
	
  # computing predictions for the full curve
	predictions <- shapes_to_fit %*% t(beta_samples)
  
	# discard first 100
	return(list(
		predictions = predictions[,(n_burnin + 1):(n_simulations + n_burnin)],
		beta_samples = beta_samples[(n_burnin + 1):(n_simulations + n_burnin),]
    ))
}

# function that receives coefficients (in normalised shifted Legendre basis) and applies Gibbs Sampling
# assumes any PCA or curve selection was done beforehand 
# and the coefficients (coefs) are the final ones to use
do_Gibbs_1D <- function(t, y, coefs, prediction_interval,
                        beta0, Sigma_0, nu0, sigmasq_0,
                        first_beta, first_sigma, n_simulations = 5000){
  
  # building basis (with normalised shifted Legendre polynomials)
  X_poly <- build_shifted_legendre_basis(t, (ncol(coefs)-1))
  
  # constructing the design matrix (evaluating the polynomials)
  X_full <- t((coefs) %*% t(X_poly))
  
  # passing things along to the inner function
  return (Gibbs_core(t, y, X_full, prediction_interval,
                     beta0, Sigma_0, nu0, sigmasq_0, 
                     first_beta, first_sigma, n_simulations))
  
}


# function that receives coefficients (in normalised shifted Legendre basis)
# for X and Y and applies Gibbs sampling
# assumes any PCA or curve selection
# was done beforehand and the coefficients (coefs) are the final ones to use.
do_Gibbs_2D <- function(t, x, y, coefs_X, coefs_Y, prediction_interval,
                        beta0, Sigma_0, nu0, sigmasq_0,
                        first_beta, first_sigma, n_simulations = 5000){
  
  # building basis (with normalised shifted Legendre polynomials)
  X_poly <- build_shifted_legendre_basis(t, (ncol(coefs_X)-1))
  Y_poly <- build_shifted_legendre_basis(t, (ncol(coefs_Y)-1))
  
  
  # constructing the design matrix (evaluating the polynomials)
  X_full <- t((coefs_X) %*% t(X_poly))
  Y_full <- t((coefs_Y) %*% t(Y_poly))
  
  # passing things along to the inner function
  return (Gibbs_core(c(t, t), c(x, y), rbind(X_full, Y_full), prediction_interval,
                     beta0, Sigma_0, nu0, sigmasq_0, 
                     first_beta, first_sigma, n_simulations))
  
}


# function that receives raw coefficients (in normalised shifted Legendre basis),
# performs curve selection, and applies Gibbs Sampling
Gibbs_1d_without_PCA <- function(t, y, full_coefs, prediction_interval,
                                 nu0 = default_nu0, sigmasq_0 = default_sigmasq_0,
                                 first_sigma = default_first_sigma,
                                 n_simulations = 5000, curve_select_number = -1) {
  
  # performing curve selection
  coefs <- curve_selection(t, y, prediction_interval, full_coefs, curve_select_number)
  
  
  # passing things along to the inner function
  return (do_Gibbs_1D(t, y, coefs, prediction_interval,
                      rep(0.5, nrow(coefs)), diag(1, nrow(coefs)), nu0, sigmasq_0,
                      rep(1/nrow(coefs), nrow(coefs)), first_sigma, n_simulations))
}

# function that receives raw coefficients (in normalised shifted Legendre basis),
# performs PCA, and applies Gibbs Sampling
Gibbs_1d_PCA <- function(t, y, coefs, prediction_interval,
                         nu0 = default_nu0, sigmasq_0 = default_sigmasq_0,
                         first_sigma = default_first_sigma,
                         n_simulations = 5000) {
  
  # building basis (with normalised shifted Legendre polynomials)
  X_poly <- build_shifted_legendre_basis(t, (ncol(coefs)-1))
  
  # finding average curve (P_0)
  P_0_cs <- colMeans(coefs)
  
  # constructing the design matrix (evaluating the polynomials)
  P_0_real <- X_poly %*% P_0_cs
  
  # new response: original response - P_0
  to_fit <- y - P_0_real
  
  # pca
  pca <- prcomp((coefs), center = TRUE, scale = TRUE)
  
  # taking eigenvectors
  new_coefs <- pca$rotation
  
  # Gibbs Sampling procedure
  res <- do_Gibbs_1D(t, to_fit, new_coefs, prediction_interval,
                     rep(0, nrow(new_coefs)), diag(1, nrow(new_coefs)), nu0, sigmasq_0,
                     rep(0, nrow(new_coefs)), first_sigma, n_simulations)
  
  # adding back P_0
  res$predictions <- res$predictions + as.vector(P_0_real)
  
  return (res)
}

# function that receives raw coefficients (in normalised shifted Legendre basis) for X and Y,
# performs curve selection, and applies Gibbs Sampling
Gibbs_2d_without_PCA <- function(t, x, y, coefs_X_full, coefs_Y_full, prediction_interval,
                                 nu0 = default_nu0, sigmasq_0 = default_sigmasq_0,
                                 first_sigma = default_first_sigma, 
                                 n_simulations = 5000, curve_select_number = -1) {
  
  suggested_num_coefs <- curve_select_number
  
  # if the number of curves is not specified,
  # it selects the right number corresponding to the
  # dimensionality of the polynomial space
  if (curve_select_number <= 0) {
    suggested_num_coefs <- ncol(coefs_X_full) + ncol(coefs_Y_full)
  }
  
  # selecting curves for both x and y
  coefs_X <- curve_selection(t, x, prediction_interval, coefs_X_full, suggested_num_coefs)
  coefs_Y <- curve_selection(t, y, prediction_interval, coefs_Y_full, suggested_num_coefs)
  
  # passing things along to the inner function
  return (do_Gibbs_2D(t, x, y, coefs_X, coefs_Y, prediction_interval,
                      rep(0.5, nrow(coefs_X)), diag(1, nrow(coefs_X)), nu0, sigmasq_0,
                      rep(1/nrow(coefs_X), nrow(coefs_X)), first_sigma, n_simulations))
}


# function that receives raw coefficients (in normalised shifted Legendre basis) for both X and Y,
# performs PCA, and applies Gibbs Sampling
Gibbs_2d_PCA <- function(t, x, y, coefs_X, coefs_Y, prediction_interval,
                         nu0 = default_nu0, sigmasq_0 = default_sigmasq_0,
                         first_sigma = default_first_sigma, n_simulations = 5000) {
  
  # building basis (with normalised shifted Legendre polynomials)
  X_poly <- build_shifted_legendre_basis(t, (ncol(coefs_X)-1))
  Y_poly <- build_shifted_legendre_basis(t, (ncol(coefs_Y)-1))

  # finding average curve (P_0)
  P_0_cs_X <- colMeans(coefs_X)
  P_0_cs_Y <- colMeans(coefs_Y)
  
  # constructing the design matrix (evaluating the polynomials)
  P_0_x <- as.vector(X_poly %*% P_0_cs_X)
  P_0_y <- as.vector(Y_poly %*% P_0_cs_Y)
  
  P_0_full <- c(P_0_x, P_0_y)
  
  # new response: original response - P_0
  to_fit_x <- x - P_0_x
  to_fit_y <- y - P_0_y
  
  coefs_XY <- cbind(coefs_X, coefs_Y)
  
  # pca
  pca <- prcomp((coefs_XY), center = TRUE, scale = TRUE)
  
  # taking eigenvectors
  new_coefs <- pca$rotation
  
  # Gibbs Sampling procedure
  res <- do_Gibbs_2D(t, to_fit_x, to_fit_y, 
                     new_coefs[,1:ncol(coefs_X)], 
                     new_coefs[,((ncol(coefs_X) + 1):ncol(new_coefs))],
                     prediction_interval,
                     rep(0, nrow(new_coefs)), diag(1, nrow(new_coefs)), nu0, sigmasq_0,
                     rep(0, nrow(new_coefs)), first_sigma, n_simulations)
  
  # adding back P_0
  res$predictions <- res$predictions + P_0_full
  
  return (res)
}


  
```

## Setting up Data Structures

### Computing Coefficients of Polynomials (in normalised shifted Legendre basis)

```{r}
# DATA STRUCTURES

# setting degrees for x and y
degree_x <- 5
degree_y <- 4

# computing shifted Legendre basis coefficients for speakers 1 and 2
coefs_X_speaker1 <- compute_betas_shifted_legendre(df1, "x", degree_x) # x curves
coefs_Y_speaker1 <- compute_betas_shifted_legendre(df1, "y", degree_y) # y curves
coefs_X_speaker2 <- compute_betas_shifted_legendre(df2, "x", degree_x) # x curves
coefs_Y_speaker2 <- compute_betas_shifted_legendre(df2, "y", degree_y) # y curves

# combining x and y curve coefficients
coefs_X <- rbind(coefs_X_speaker1, coefs_X_speaker2) # x curves
coefs_Y <- rbind(coefs_Y_speaker1, coefs_Y_speaker2) # y curves
```

### Determining Missing Regions

```{r}
# list of tongue shapes with missing values (first two shapes of both speakers)
curves_to_check <- list(c(1, 1), c(1, 2), c(2, 1), c(2, 2))

# list of missing regions; each region is described by (tmin, tmax), where t is coordinate along the curve
missing_regions <- list()

for (c in curves_to_check){
  
  # selecting data for a given speaker and curve
  this_curve <- subset(df, df$speaker == c[1] & df$shape == c[2])
  
  # determining the missing section
  this_missing <- is.na(this_curve$x) | is.na(this_curve$y)
  
  # calculating the beginning and the end of the missing sections through the indices t (standardised)
  this_min <- min(this_curve$t_standardised[this_missing]) - 1/(nrow(this_curve) + 1)
  this_max <- max(this_curve$t_standardised[this_missing]) + 1/(nrow(this_curve) + 1)
  
  missing_regions[[length(missing_regions) + 1]] <- c(this_min, this_max)
}
```


## Descriptive Overview

The following subsection contains the figures presented in the poster (excluding results which are given later).

### Tongue Shapes of Both Speakers (Figures 1 and 2)

```{r}
# 2D curves of speaker 1
ggplot(df1, aes(x = x, y = y)) +
  geom_point(size = 1) +                       
  facet_wrap(~shape, ncol = 10) +      
  coord_fixed() + labs(x = NULL, y = NULL) +                
  theme_bw()

# 2D curves of speaker 2
ggplot(df2, aes(x = x, y = y)) +
  geom_point(size = 1) +                       
  facet_wrap(~shape, ncol = 10) +      
  coord_fixed() +  labs(x = NULL, y = NULL) +  
  theme_bw()

#ggsave("speaker1.png", plot = p1, width = 27, height = 4, units = "cm", dpi = 300)
#ggsave("speaker2.png", plot = p2, width = 27, height = 4, units = "cm", dpi = 300)
```

### Curves of x-coordinates (Figure 3)

```{r}
# grid of points between 0 and 1
x_grid <- seq(0, 1, length.out = 100)
degree <- ncol(coefs_X)

# building the shifted legendre basis for the points
XX_poly <- build_shifted_legendre_basis(x_grid, degree - 1)

# evaluating points
curves <- as.data.frame(XX_poly %*% t(coefs_X))
curves$x <- x_grid

# transforming into a long format
df_long <- curves %>%
  pivot_longer(cols = -x, names_to = "curve", values_to = "y") %>%
  mutate(
    curve       = as.integer(gsub("V", "", curve)),     # number of the curve
    color_group = ((curve - 1) %% 8) + 1,               # determining colour
    lty_group   = ifelse(curve <= 8, "solid", "dashed") # style
  )

# plotting
base_colors <- rainbow(8)
p1 <- ggplot(df_long, aes(x = x, y = y, group = curve)) +
  geom_line(aes(color = factor(color_group), linetype = lty_group), linewidth = 0.5) +
  scale_color_manual(values = setNames(base_colors, as.character(1:8)), guide = "none") +
  scale_linetype_manual(values = c("solid" = 1, "dashed" = 2), guide = "none") +
  labs(x = expression(tau), y = "x") + theme_bw()

################################################################################
# save
#ggsave("curves_x.png", plot = p1, width = 9.75, height = 6.7, units = "cm", dpi = 300)

```

### Curves of y Coordinates (Figure 3)

```{r}
# grid of points between 0 and 1
y_grid <- seq(0, 1, length.out = 100)
degree <- ncol(coefs_Y)

# building the shifted legendre basis for the points
Y_poly <- build_shifted_legendre_basis(y_grid, degree - 1)

# evaluating points
curves <- as.data.frame(Y_poly %*% t(coefs_Y))
curves$x <- y_grid

# transforming into a long format
df_long <- curves %>%
  pivot_longer(cols = -x, names_to = "curve", values_to = "y") %>%
  mutate(
    curve       = as.integer(gsub("V", "", curve)),     # number of the curve
    color_group = ((curve - 1) %% 8) + 1,               # determining colour
    lty_group   = ifelse(curve <= 8, "solid", "dashed") # style
  )

# plotting
base_colors <- rainbow(8)
p1 <- ggplot(df_long, aes(x = x, y = y, group = curve)) +
  geom_line(aes(color = factor(color_group), linetype = lty_group), linewidth = 0.5) +
  scale_color_manual(values = setNames(base_colors, as.character(1:8)), guide = "none") +
  scale_linetype_manual(values = c("solid" = 1, "dashed" = 2), guide = "none") +
  labs(x = expression(tau), y = "y") + theme_bw()

################################################################################
# save
#ggsave("curves_y.png", plot = p1, width = 9.75, height = 6.7, units = "cm", dpi = 300)

```


## Results

The following subsection contains results of all four methods.

### Preliminary Step

```{r}

# finding row indices corresponding to missing regions (needed for plotting)
curves_to_check <- list(c(1, 1), c(1, 2), c(2, 1), c(2, 2))

regions <- list()
i <- 1

for (c in curves_to_check){
  
  # selecting data for a given speaker and curve
  this_curve <- subset(df, df$speaker == c[1] & df$shape == c[2])
  
  # determining the missing interval
  regions[[i]] <- which(
             this_curve$t_standardised > missing_regions[[i]][1] & 
             this_curve$t_standardised < missing_regions[[i]][2])
  
  i <- i + 1
  
  }

# storing results:
# region 1 corresponds to speaker 1, curve 1
# region 2 corresponds to speaker 1, curve 2
# region 3 corresponds to speaker 2, curve 1
# region 4 corresponds to speaker 2, curve 2
region_1 <- regions[[1]]
region_2 <- regions[[2]]
region_3 <- regions[[3]]
region_4 <- regions[[4]]
```


### Speaker 1, Curve 1

```{r}

set.seed(123)
################################################################################

# marginal modelling; with PCA
speaker_1_shape_1_x_PCA <- Gibbs_1d_PCA(subset(df1, shape == 1)$t_standardised, subset(df1, shape == 1)$x, coefs_X, missing_regions[[1]])
speaker_1_shape_1_y_PCA <- Gibbs_1d_PCA(subset(df1, shape == 1)$t_standardised, subset(df1, shape == 1)$y, coefs_Y, missing_regions[[1]])

# x curve
summarise_results_1d(speaker_1_shape_1_x_PCA$predictions, subset(df1, shape == 1)$t_standardised, subset(df1, shape == 1)$x, region_1, "x")
# y curve
summarise_results_1d(speaker_1_shape_1_y_PCA$predictions, subset(df1, shape == 1)$t_standardised, subset(df1, shape == 1)$y, region_1, "y")
# together (x,y)
p1 <- summarise_results_2d(speaker_1_shape_1_x_PCA$predictions, speaker_1_shape_1_y_PCA$predictions, subset(df1, shape == 1)$x, subset(df1, shape == 1)$y, region_1, c(10,50), c(35,65))

################################################################################

# marginal modelling; without PCA
speaker_1_shape_1_x_without_PCA <- Gibbs_1d_without_PCA(subset(df1, shape == 1)$t_standardised, subset(df1, shape == 1)$x, coefs_X, missing_regions[[1]])
speaker_1_shape_1_y_without_PCA <- Gibbs_1d_without_PCA(subset(df1, shape == 1)$t_standardised, subset(df1, shape == 1)$y, coefs_Y, missing_regions[[1]])

# x curve
summarise_results_1d(speaker_1_shape_1_x_without_PCA$predictions, subset(df1, shape == 1)$t_standardised, subset(df1, shape == 1)$x, region_1, "x")
# y curve
summarise_results_1d(speaker_1_shape_1_y_without_PCA$predictions, subset(df1, shape == 1)$t_standardised, subset(df1, shape == 1)$y, region_1,  "y")
# together (x,y)
p2 <- summarise_results_2d(speaker_1_shape_1_x_without_PCA$predictions, speaker_1_shape_1_y_without_PCA$predictions, subset(df1, shape == 1)$x, subset(df1, shape == 1)$y, region_1, c(10,50), c(35,65))

################################################################################

# joint modelling; with PCA
speaker1_curve1_PCA <- Gibbs_2d_PCA(subset(df1, shape == 1)$t_standardised, subset(df1, shape == 1)$x, subset(df1, shape == 1)$y, coefs_X, coefs_Y, missing_regions[[1]])

# together (x,y)
p3 <- summarise_results_2d(speaker1_curve1_PCA$predictions[1:65,], speaker1_curve1_PCA$predictions[66:130,], subset(df1, shape == 1)$x, subset(df1, shape == 1)$y, region_1, c(10,50), c(35,65))

################################################################################

# joint modelling; without PCA
speaker1_curve1_without_PCA <- Gibbs_2d_without_PCA(subset(df1, shape == 1)$t_standardised, subset(df1, shape == 1)$x, subset(df1, shape == 1)$y, coefs_X, coefs_Y, missing_regions[[1]], 5000)

# together (x,y)
p4 <- summarise_results_2d(speaker1_curve1_without_PCA$predictions[1:65,], speaker1_curve1_without_PCA$predictions[66:130,], subset(df1, shape == 1)$x, subset(df1, shape == 1)$y, region_1, c(10,50), c(35,65))

################################################################################
# save
ggsave("speaker1_curve1_1d_pca.png", plot = p1, width = 10.7, height = 7.7, units = "cm", dpi = 300)
ggsave("speaker1_curve1_1d_withoutpca.png", plot = p2, width = 10.7, height = 7.7, units = "cm", dpi = 300)
ggsave("speaker1_curve1_2d_pca.png", plot = p3, width = 10.7, height = 7.7, units = "cm", dpi = 300)
ggsave("speaker1_curve1_2d_withoutpca.png", plot = p4, width = 10.7, height = 7.7, units = "cm", dpi = 300)

```

### Speaker 1, Curve 2

```{r}
set.seed(123)
################################################################################

# marginal modelling; with PCA
speaker_1_shape_2_x_PCA <- Gibbs_1d_PCA(subset(df1, shape == 2)$t_standardised, subset(df1, shape == 2)$x, coefs_X, missing_regions[[2]])
speaker_1_shape_2_y_PCA <- Gibbs_1d_PCA(subset(df1, shape == 2)$t_standardised, subset(df1, shape == 2)$y, coefs_Y, missing_regions[[2]])

# x curve
summarise_results_1d(speaker_1_shape_2_x_PCA$predictions, subset(df1, shape == 2)$t_standardised, subset(df1, shape == 2)$x, region_2, "x")
# y curve
summarise_results_1d(speaker_1_shape_2_y_PCA$predictions, 1:65, subset(df1, shape == 2)$y, region_2, "y")
# together (x,y)
p1 <- summarise_results_2d(speaker_1_shape_2_x_PCA$predictions, speaker_1_shape_2_y_PCA$predictions, subset(df1, shape == 2)$x, subset(df1, shape == 2)$y, region_2, c(15,47), c(35,65))

################################################################################

# marginal modelling; without PCA
speaker_1_shape_2_x_without_PCA <- Gibbs_1d_without_PCA(subset(df1, shape == 2)$t_standardised, subset(df1, shape == 2)$x, coefs_X, missing_regions[[2]])
speaker_1_shape_2_y_without_PCA <- Gibbs_1d_without_PCA(subset(df1, shape == 2)$t_standardised, subset(df1, shape == 2)$y, coefs_Y, missing_regions[[2]])

# x curve
summarise_results_1d(speaker_1_shape_2_x_without_PCA$predictions, subset(df1, shape == 2)$t_standardised, subset(df1, shape == 2)$x, region_2, "x")
# y curve
summarise_results_1d(speaker_1_shape_2_y_without_PCA$predictions, subset(df1, shape == 2)$t_standardised, subset(df1, shape == 2)$y, region_2, "y")
# together (x,y)
p2 <- summarise_results_2d(speaker_1_shape_2_x_without_PCA$predictions, speaker_1_shape_2_y_without_PCA$predictions, subset(df1, shape == 2)$x, subset(df1, shape == 2)$y, region_2, c(15,47), c(35,65))

################################################################################

# joint modelling; with PCA
speaker1_curve2_PCA <- Gibbs_2d_PCA(subset(df1, shape == 2)$t_standardised, subset(df1, shape == 2)$x, subset(df1, shape == 2)$y, coefs_X, coefs_Y, missing_regions[[2]])

# together (x,y)
p3 <- summarise_results_2d(speaker1_curve2_PCA$predictions[1:65,], speaker1_curve2_PCA$predictions[66:130,], subset(df1, shape == 2)$x, subset(df1, shape == 2)$y, region_2, c(15,47), c(35,65))

################################################################################

# joint modelling; without PCA
speaker1_curve2_without_PCA <- Gibbs_2d_without_PCA(subset(df1, shape == 2)$t_standardised, subset(df1, shape == 2)$x, subset(df1, shape == 2)$y, coefs_X, coefs_Y, missing_regions[[2]])

# together (x,y)
p4 <- summarise_results_2d(speaker1_curve2_without_PCA$predictions[1:65,], speaker1_curve2_without_PCA$predictions[66:130,], subset(df1, shape == 2)$x, subset(df1, shape == 2)$y, region_2, c(15,47), c(35,65))

################################################################################
# save
ggsave("speaker1_curve2_1d_pca.png", plot = p1, width = 10.7, height = 7.7, units = "cm", dpi = 300)
ggsave("speaker1_curve2_1d_withoutpca.png", plot = p2, width = 10.7, height = 7.7, units = "cm", dpi = 300)
ggsave("speaker1_curve2_2d_pca.png", plot = p3, width = 10.7, height = 7.7, units = "cm", dpi = 300)
ggsave("speaker1_curve2_2d_withoutpca.png", plot = p4, width = 10.7, height = 7.7, units = "cm", dpi = 300)

```


###  Speaker 2, Curve 1

```{r}
set.seed(123)
################################################################################

# marginal modelling; with PCA
speaker_2_shape_1_x_PCA <- Gibbs_1d_PCA(subset(df2, shape == 1)$t_standardised, subset(df2, shape == 1)$x, coefs_X, missing_regions[[3]])
speaker_2_shape_1_y_PCA <- Gibbs_1d_PCA(subset(df2, shape == 1)$t_standardised, subset(df2, shape == 1)$y, coefs_Y, missing_regions[[3]])

# x curve
summarise_results_1d(speaker_2_shape_1_x_PCA$predictions, subset(df2, shape == 1)$t_standardised, subset(df2, shape == 1)$x, region_3, "x")
# y curve
summarise_results_1d(speaker_2_shape_1_y_PCA$predictions, subset(df2, shape == 1)$t_standardised, subset(df2, shape == 1)$y, region_3, "y")
# together (x,y)
p1 <- summarise_results_2d(speaker_2_shape_1_x_PCA$predictions, speaker_2_shape_1_y_PCA$predictions, subset(df2, shape == 1)$x, subset(df2, shape == 1)$y, region_3, c(12,50), c(30,65))

################################################################################

# marginal modelling; without PCA
speaker_2_shape_1_x_without_PCA <- Gibbs_1d_without_PCA(subset(df2, shape == 1)$t_standardised, subset(df2, shape == 1)$x, coefs_X, missing_regions[[3]])
speaker_2_shape_1_y_without_PCA <- Gibbs_1d_without_PCA(subset(df2, shape == 1)$t_standardised, subset(df2, shape == 1)$y, coefs_Y, missing_regions[[3]])

# x curve
summarise_results_1d(speaker_2_shape_1_x_without_PCA$predictions, subset(df2, shape == 1)$t_standardised, subset(df2, shape == 1)$x, region_3, "x")
# y curve
summarise_results_1d(speaker_2_shape_1_y_without_PCA$predictions, subset(df2, shape == 1)$t_standardised, subset(df2, shape == 1)$y, region_3, "y")
# together (x,y)
p2 <- summarise_results_2d(speaker_2_shape_1_x_without_PCA$predictions, speaker_2_shape_1_y_without_PCA$predictions, subset(df2, shape == 1)$x, subset(df2, shape == 1)$y, region_3, c(12,50), c(30,65))

################################################################################

# joint modelling; with PCA
speaker2_curve1_PCA <- Gibbs_2d_PCA(subset(df2, shape == 1)$t_standardised, subset(df2, shape == 1)$x, subset(df2, shape == 1)$y, coefs_X, coefs_Y, missing_regions[[3]])

# together (x,y)
p3 <- summarise_results_2d(speaker2_curve1_PCA$predictions[1:45,], speaker2_curve1_PCA$predictions[46:90,], subset(df2, shape == 1)$x, subset(df2, shape == 1)$y, region_3, c(12,50), c(30,65))

################################################################################

# joint modelling; without PCA
speaker2_curve1_without_PCA <- Gibbs_2d_without_PCA(subset(df2, shape == 1)$t_standardised, subset(df2, shape == 1)$x, subset(df2, shape == 1)$y, coefs_X, coefs_Y, missing_regions[[3]])

# together (x,y)
p4 <- summarise_results_2d(speaker2_curve1_without_PCA$predictions[1:45,], speaker2_curve1_without_PCA$predictions[46:90,], subset(df2, shape == 1)$x, subset(df2, shape == 1)$y, region_3, c(12,50), c(30,65))

################################################################################
# save
ggsave("speaker2_curve1_1d_pca.png", plot = p1, width = 10.7, height = 7.7, units = "cm", dpi = 300)
ggsave("speaker2_curve1_1d_withoutpca.png", plot = p2, width = 10.7, height = 7.7, units = "cm", dpi = 300)
ggsave("speaker2_curve1_2d_pca.png", plot = p3, width = 10.7, height = 7.7, units = "cm", dpi = 300)
ggsave("speaker2_curve1_2d_withoutpca.png", plot = p4, width = 10.7, height = 7.7, units = "cm", dpi = 300)


```


### Speaker 2, Curve 2

```{r}
set.seed(123)
################################################################################

# marginal modelling; with PCA
speaker_2_shape_2_x_PCA <- Gibbs_1d_PCA(subset(df2, shape == 2)$t_standardised, subset(df2, shape == 2)$x, coefs_X, missing_regions[[4]])
speaker_2_shape_2_y_PCA <- Gibbs_1d_PCA(subset(df2, shape == 2)$t_standardised, subset(df2, shape == 2)$y, coefs_Y, missing_regions[[4]])

# x curve
summarise_results_1d(speaker_2_shape_2_x_PCA$predictions, subset(df2, shape == 2)$t_standardised, subset(df2, shape == 2)$x, region_4, "x")
# y curve
summarise_results_1d(speaker_2_shape_2_y_PCA$predictions, subset(df2, shape == 2)$t_standardised, subset(df2, shape == 2)$y, region_4, "y")
# together (x,y)
p1 <- summarise_results_2d(speaker_2_shape_2_x_PCA$predictions, speaker_2_shape_2_y_PCA$predictions, subset(df2, shape == 2)$x, subset(df2, shape == 2)$y, region_4, c(13,40), c(47,65))

################################################################################

# marginal modelling; without PCA
speaker_2_shape_2_x_without_PCA <- Gibbs_1d_without_PCA(subset(df2, shape == 2)$t_standardised, subset(df2, shape == 2)$x, coefs_X, missing_regions[[4]])
speaker_2_shape_2_y_without_PCA <- Gibbs_1d_without_PCA(subset(df2, shape == 2)$t_standardised, subset(df2, shape == 2)$y, coefs_Y, missing_regions[[4]])

# x curve
summarise_results_1d(speaker_2_shape_2_x_without_PCA$predictions, subset(df2, shape == 2)$t_standardised, subset(df2, shape == 2)$x, region_4, "x")
# y curve
summarise_results_1d(speaker_2_shape_2_y_without_PCA$predictions, subset(df2, shape == 2)$t_standardised, subset(df2, shape == 2)$y, region_4, "y")
# together (x,y)
p2 <- summarise_results_2d(speaker_2_shape_2_x_without_PCA$predictions, speaker_2_shape_2_y_without_PCA$predictions, subset(df2, shape == 2)$x, subset(df2, shape == 2)$y, region_4, c(13,40), c(47,65))

################################################################################

# joint modelling; with PCA
speaker2_curve2_PCA <- Gibbs_2d_PCA(subset(df2, shape == 2)$t_standardised, subset(df2, shape == 2)$x, subset(df2, shape == 2)$y, coefs_X, coefs_Y, missing_regions[[4]])

# together (x,y)
p3 <- summarise_results_2d(speaker2_curve2_PCA$predictions[1:45,], speaker2_curve2_PCA$predictions[46:90,], subset(df2, shape == 2)$x, subset(df2, shape == 2)$y, region_4, c(13,40), c(47,65))

################################################################################

# joint modelling; without PCA
speaker2_curve2_without_PCA <- Gibbs_2d_without_PCA(subset(df2, shape == 2)$t_standardised, subset(df2, shape == 2)$x, subset(df2, shape == 2)$y, coefs_X, coefs_Y, missing_regions[[4]])

# together (x,y)
p4 <- summarise_results_2d(speaker2_curve2_without_PCA$predictions[1:45,], speaker2_curve2_without_PCA$predictions[46:90,], subset(df2, shape == 2)$x, subset(df2, shape == 2)$y, region_4, c(13,40), c(47,65))

################################################################################
# save
ggsave("speaker2_curve2_1d_pca.png", plot = p1, width = 10.7, height = 7.7, units = "cm", dpi = 300)
ggsave("speaker2_curve2_1d_withoutpca.png", plot = p2, width = 10.7, height = 7.7, units = "cm", dpi = 300)
ggsave("speaker2_curve2_2d_pca.png", plot = p3, width = 10.7, height = 7.7, units = "cm", dpi = 300)
ggsave("speaker2_curve2_2d_withoutpca.png", plot = p4, width = 10.7, height = 7.7, units = "cm", dpi = 300)



```


### Comparison of PCA vs non-PCA in Marginal Modelling (Figure 5)

```{r}
#model_order <- c("No PCA", "PCA")
model_order <- c("PCA", "No PCA")


# speaker 1; curve 1; x curve
p1 <- plot_1d_models(
      model_list = list(
     "No PCA" = speaker_1_shape_1_x_without_PCA,
     "PCA" = speaker_1_shape_1_x_PCA), subset(df1, shape == 1)$t_standardised, 
     subset(df1, shape == 1)$x, region_1,
     colors = c("No PCA" = "red", "PCA" = "blue"), response = "x", model_order)

# speaker 1; curve 1; y curve
p2 <- plot_1d_models(
     model_list = list(
    "No PCA" = speaker_1_shape_1_y_without_PCA,
    "PCA" = speaker_1_shape_1_y_PCA), subset(df1, shape == 1)$t_standardised, 
     subset(df1, shape == 1)$y, region_1,
     colors = c("No PCA" = "red", "PCA" = "blue"), response = "y", model_order)

################################################################################
# save
ggsave("marginal_x.png", plot = p1, width = 9.75, height = 6.7, units = "cm", dpi = 300)
ggsave("marginal_y.png", plot = p2, width = 9.75, height = 6.7, units = "cm", dpi = 300)


```


### Model Comparison


The following subsection contains code for model comparison and all relevant parts that were used for summarising the results, as presented on the poster.


#### Finding Average Error for Each Region and Model (Figure 7)

```{r}
set.seed(123)

################################################################################
# helper functions 

# function that computes the sum of absolute errors between predictions 
# and true points over the interval of interest
calculate_SSR <- function(predictions, true, to_be_predicted){
  
  best_estimate <- apply(predictions, 1, quantile, probs = 0.5)[to_be_predicted]
  true <- true[to_be_predicted]
  SSR <- sum(abs(best_estimate - true))
  
  return((SSR))
}

# function that computes errors for all four methods for a given curve and missing region
error <- function(coefs_X, coefs_Y, t_standardised, true_x, true_y, missing_region){
  
  n <- length(true_x)
  
  # with PCA
  x_PCA <- Gibbs_1d_PCA(t_standardised, true_x, coefs_X, missing_region)
  y_PCA <- Gibbs_1d_PCA(t_standardised, true_y, coefs_Y, missing_region)
  xy_2d_PCA <- Gibbs_2d_PCA(t_standardised, true_x, true_y, coefs_X, coefs_Y, missing_region)
  
  # without PCA
  x_no_PCA <- Gibbs_1d_without_PCA(t_standardised, true_x, coefs_X, missing_region)
  y_no_PCA <- Gibbs_1d_without_PCA(t_standardised, true_y, coefs_Y, missing_region)
  xy_2d_no_PCA <- Gibbs_2d_without_PCA(t_standardised, true_x, true_y, coefs_X, coefs_Y, missing_region)
  
  # determining points over which the error is computed
  to_be_predicted <- (t_standardised > missing_region[1]) & (t_standardised < missing_region[2])
  
  # computing errors for each method; x and y parts are done separately
  m1 <- calculate_SSR(x_PCA$predictions, true_x, to_be_predicted) + calculate_SSR(y_PCA$predictions, true_y, to_be_predicted)
  
  m2 <- calculate_SSR(x_no_PCA$predictions, true_x, to_be_predicted) + calculate_SSR(y_no_PCA$predictions, true_y, to_be_predicted)
  
  m3 <- calculate_SSR(xy_2d_PCA$predictions[1:n,], true_x, to_be_predicted) + calculate_SSR(xy_2d_PCA$predictions[(n+1):(2*n),], true_y, to_be_predicted)
  
   m4 <- calculate_SSR(xy_2d_no_PCA$predictions[1:n,], true_x, to_be_predicted) + calculate_SSR(xy_2d_no_PCA$predictions[(n+1):(2*n),], true_y, to_be_predicted)
   
   return(c((m1), (m2), (m3), (m4)))
}

################################################################################
################################################################################

# all pairs of (shape, speaker) used in model comparison (all fully observed tongue shapes)
curves_to_fit <- expand.grid(3:10, 1:2)
n_coefs_per_speaker <- 8
n_ignored_curves <- 2
n_methods <- 4

# initialising data structure to store results
diagnostic_results <- matrix(NA, nrow = nrow(curves_to_fit), ncol = n_methods * length(missing_regions))

for (i in 1:nrow(curves_to_fit)) {
  
  this_shape <- curves_to_fit[i, 1]
  this_speaker <- curves_to_fit[i, 2]
  
  this_coef <- n_coefs_per_speaker * (this_speaker - 1) + (this_shape - n_ignored_curves)
  
  # indices such that the coefficients of a given shape used for error computation are excluded 
  # in order to avoid leakage
  selected_coefs <- (1:nrow(coefs_X))[-this_coef]
  
  # extracting true data for a given shape
  this_curve <- subset(df, df$speaker == this_speaker & df$shape == this_shape)
  
  # computing errors
  for (j in 1:length(missing_regions)) {
  
    
    results <- error(coefs_X[selected_coefs,], coefs_Y[selected_coefs,],
                     this_curve$t_standardised, this_curve$x, this_curve$y,
                     missing_regions[[j]])
    
    
    diagnostic_results[i, (1 + (j-1) * length(results)):(j* length(results))] <- results
  }
}


# in order to have comparable values between speaker 1 and 2, dividing by the number of points
normalised_res <- rbind(diagnostic_results[1:8,] / 65 , diagnostic_results[9:16,] / 45)

################################################################################
# BUILDING TABLE OF RESULTS

# mapping of methods based on original table
col_map <- list(
  "1" = c(1,5,9,13),
  "2" = c(2,6,10,14),
  "3" = c(3,7,11,15),
  "4" = c(4,8,12,16))

models <- 1:4
methods_table <- list()

for (m in models) {
  
  # selecting model
  columns <- col_map[[m]]
  # extracting corresponding chunk
  this_res <- normalised_res[, columns]
  # taking transpose: rows = regions, columns = curves
  this_res <- t(this_res)
  
  # convert to data frame
  df_this_res <- as.data.frame(this_res)
  
  # average per region for each model
  # (taking all points into account)
  df_this_res$region_average <- (colSums(diagnostic_results[, columns])/(8 * 65 + 8 * 45))
  
  # label for region
  df_this_res$missing_region <- rep(1:4)
  
  # label for model
  model_names <- c("Marginal PCA", "Marginal without PCA", "Joint PCA", "Joint without PCA")
  df_this_res$model <- model_names[m]
  
  methods_table[[m]] <- df_this_res
}

# combining all rows
results_per_method <- do.call(rbind, methods_table)
# computing mean per model (across all regions)
model_means <- aggregate(region_average ~ model, results_per_method, mean)
# correct order (because aggregate does not keep the right order)
correct_order <- c("Marginal PCA", "Marginal without PCA", "Joint PCA", "Joint without PCA")
model_means <- model_means[match(correct_order, model_means$model), ]

# adding mean to dataframe
results_per_method$model_average <- rep(model_means$region_average, each = 4)
# rounding
numeric_columns <- sapply(results_per_method, is.numeric)
results_per_method[numeric_columns] <- round(results_per_method[numeric_columns], 2)
# results (move model and missing region to the beginning)
final_table <- results_per_method[, c("model", "missing_region", setdiff(names(results_per_method), c("model", "missing_region")))]

# save
#write.csv(final_table, "final.csv", row.names = FALSE)

################################################################################
# CREATING A PLOT

final_table$missing_region <- as.factor(final_table$missing_region)
# to ensure right order
final_table$model <- factor(final_table$model,
                     levels = c("Marginal without PCA", "Joint without PCA", "Joint PCA", "Marginal PCA"))

# plot (log scale)
final <- ggplot(final_table, aes(x = model, y = region_average, fill = missing_region)) +
geom_errorbar(aes(ymin = model_average, ymax = model_average), 
              linewidth = 0.8, width = 0.9, position = "identity", linetype = "dashed") +
 geom_col(position = position_dodge(width = 0.85), width = 0.75, alpha = 1) +
  scale_fill_manual(values = c("darkred", "darkblue", "darkgreen", "darkgoldenrod"))+
  labs(x = "Model", y = "Absolute Error", fill = "Missing Region")  + 
  theme_bw() + 
  theme(axis.text.x = element_text(size = 13), 
        legend.text = element_text(size = 13),
        legend.title = element_text(size = 13))

# save
ggsave("final2.png", plot = final, width = 25, height = 16, units = "cm", dpi = 300)
################################################################################

```

#### Finding Point-Wise Errors for Each Region and Model (Figure 6)

```{r}
set.seed(123)

################################################################################
# helper functions 

# function that computes the absolute errors between predictions 
# and true points over the interval of interest
calculate_pointwise_error <- function(predictions, true, to_be_predicted){
  
  best_estimate <- apply(predictions, 1, quantile, probs = 0.5)
  
  errors <- abs(best_estimate - true)
  
  return((errors[to_be_predicted]))
}

# function that computes point-wise errors for all four methods for a given curve and missing region
error2 <- function(coefs_X, coefs_Y, t_standardised,
                  true_x, true_y, missing_region){
  
  n <- length(true_x)
  
  # with PCA; marginal modelling
  x_PCA <- Gibbs_1d_PCA(t_standardised, true_x, coefs_X, missing_region)
  y_PCA <- Gibbs_1d_PCA(t_standardised, true_y, coefs_Y, missing_region)
  
  # without PCA, marginal modelling
  x_no_PCA <- Gibbs_1d_without_PCA(t_standardised, true_x, coefs_X, missing_region)
  y_no_PCA <- Gibbs_1d_without_PCA(t_standardised, true_y, coefs_Y, missing_region)
  
  # with PCA; joint modelling
  xy_2d_PCA <- Gibbs_2d_PCA(t_standardised, true_x, true_y, coefs_X, coefs_Y, missing_region)
  
  # without PCA; joint modelling
  xy_2d_no_PCA <- Gibbs_2d_without_PCA(t_standardised, true_x, true_y, coefs_X, coefs_Y, missing_region)
  
  # determining points over which the error is computed
  to_be_predicted <- (t_standardised > missing_region[1]) &
    (t_standardised < missing_region[2])
  
  # computing errors for each method; x and y parts are done separately
  m1_x <- calculate_pointwise_error(x_PCA$predictions, true_x, to_be_predicted)
  m1_y <- calculate_pointwise_error(y_PCA$predictions, true_y, to_be_predicted)
  
  m2_x <- calculate_pointwise_error(x_no_PCA$predictions, true_x, to_be_predicted)
  m2_y <- calculate_pointwise_error(y_no_PCA$predictions, true_y, to_be_predicted)
  
  m3_x <- calculate_pointwise_error(xy_2d_PCA$predictions[1:n,], true_x, to_be_predicted)
  m3_y <- calculate_pointwise_error(xy_2d_PCA$predictions[(n+1):(2*n),], true_y, to_be_predicted)
  
  m4_x <- calculate_pointwise_error(xy_2d_no_PCA$predictions[1:n,], true_x, to_be_predicted)
  m4_y <- calculate_pointwise_error(xy_2d_no_PCA$predictions[(n+1):(2*n),], true_y, to_be_predicted)
  
  # returning a list of results; x and y parts are summed element-wise
  list(
    m1 = (m1_x + m1_y),
    m2 = (m2_x + m2_y),
    m3 = (m3_x + m3_y),
    m4 = (m4_x + m4_y)
  )
}

################################################################################
################################################################################

# all pairs of (shape, speaker) used in model comparison (all fully observed tongue shapes)
curves_to_fit <- expand.grid(3:10, 1:2)
n_coefs_per_speaker <- 8
n_ignored_curves <- 2
n_methods <- 4

# initialising list to store results
diagnostic_results2 <- list()
for (i in 1:4) {
  diagnostic_results2[[i]] <- list(c(), c(), c(), c())
}



for (i in 1:nrow(curves_to_fit)) {
  
  this_shape <- curves_to_fit[i, 1]
  this_speaker <- curves_to_fit[i, 2]
  
  this_coef <- n_coefs_per_speaker * (this_speaker - 1) + (this_shape - n_ignored_curves)
  
  # indices such that the coefficients of a given shape used for error computation are excluded 
  # in order to avoid leakage
  selected_coefs <- (1:nrow(coefs_X))[-this_coef]
  
  # extracting true data for a given shape
  this_curve <- subset(df, df$speaker == this_speaker & df$shape == this_shape)
  
  # computing errors
  for (j in 1:length(missing_regions)) {
    

    results <- error2(coefs_X[selected_coefs,], coefs_Y[selected_coefs,],
                     this_curve$t_standardised, this_curve$x, this_curve$y,
                     missing_regions[[j]])
    
    
    for (k in 1:length(results)) {
      diagnostic_results2[[j]][[k]] <- c(diagnostic_results2[[j]][[k]], results[[k]])
    }
    
  }
}


################################################################################
# PLOTS

# function that plots errors for a given model
# number of bins can be specified to ensure all plots are on comparable scale in the end
summarise_point_wise_errors <- function(method, diagnostic_results2, region_names, nr_of_bins){
  
  # data for all 4 regions
  df_hist_region <- data.frame(
  value = unlist(lapply(1:4, function(j) diagnostic_results2[[j]][[method]])),
  Region = rep(region_names, times = sapply(1:4, function(j) length(diagnostic_results2[[j]][[method]]))))

  # plotting
  p <- ggplot(df_hist_region, aes(x = value, fill = Region)) +
  geom_histogram(position = "stack", bins = nr_of_bins) +
  scale_fill_manual(values = c("darkred", "darkblue", "darkgreen", "darkgoldenrod")) +
  labs(x = NULL, y = NULL)  + coord_cartesian(xlim = c(0, 48), ylim = c(0,125)) +
  scale_x_continuous(expand = c(0.01, 0)) +
  theme_bw() + theme(legend.position = "none")
  
  return(p)
  
}

# list of regions
region_names <- c("1", "2", "3", "4")

# plots
p1 <- summarise_point_wise_errors(1, diagnostic_results2, region_names, 50)
p2 <- summarise_point_wise_errors(2, diagnostic_results2, region_names, 140)
p3 <- summarise_point_wise_errors(3, diagnostic_results2, region_names, 50)
p4 <- summarise_point_wise_errors(4, diagnostic_results2, region_names, 100)



# save
ggsave("errors_pca_marginal.png", plot = p1, width = 15, height = 8, units = "cm", dpi = 300)
ggsave("errors_withoutpca_marginal.png", plot = p2, width = 15, height = 8, units = "cm", dpi = 300)
ggsave("errors_pca_joint.png", plot = p3, width = 15, height = 8, units = "cm", dpi = 300)
ggsave("errors_withoutpca_joint.png", plot = p4, width = 15, height = 8, units = "cm", dpi = 300)
```



